The Unicode Cookbook \linebreak for Linguists

### 2018-06-04 book/chapters/implementation.tex
fixes en dashes throughout

Some characters look identical, although they are encoded differently.
Unicode offers different ways of normalization (see
Section~\ref{pitfall-canonical-equivalence}), which can be invoked here as well
using the option \texttt{normalize}. NFC-normalization turns everything into the
precomposed characters, while NFD-normalization separates everything into base
characters with combining diacritics. Splitting by code points (i.e. \texttt{sep~=~""}) shows the results of these two normalizations.

>>>>

Some characters look identical, although they are encoded differently.
Unicode offers different ways of normalization (see
Section~\ref{pitfall-canonical-equivalence}), which can be invoked here as well
using the option \texttt{normalize}. NFC normalization turns everything into the
precomposed characters, while NFD normalization separates everything into base
characters with combining diacritics. Splitting by code points (i.e. \texttt{sep~=~""}) shows the results of these two normalizations.

### 2018-06-04 book/chapters/introduction.tex
fixes en dashes throughout

Writing systems arise and develop in a complex mixture of cultural,
technological and practical pressures. They tend to be highly conservative, in
that people who have learned to read and write in a specific way---however
impractical or tedious---are mostly unwilling to change their habits. Writers
tend to resist spelling reforms. In all literate societies there exists a strong
socio-political mainstream that tries to force unification of writing (for
example by strongly enforcing ``right'' from ``wrong'' writing in schools).
However, there is also a large community of users who take as many liberties in
their writing as they can get away with.

@@

In the nineteenth century, electric telegraphy lead to a next innovation in
which written language characters were encoded by signals sent through a copper
wire. Originally, \textsc{bisignal codes} were used, consisting of two different
signals. For example, Carl Friedrich Gauss in 1833 used positive and negative
current \citep[282]{Mania2008}. More famous and influential, Samuel Morse in
1836 used long and short pulses. In those bisignal codes each character from the
written language was encoded with a different number of signals (between one and
five), so two different separators are needed: one between signals and one
between characters. For example, in Morse-code there is a short pause between
signals and a long pause between characters.\footnote{Actually, Morse-code also
includes an extra long pause between words. Interestingly, it took a long time
to consider the written word boundary---using white-space---as a bona-fide
character that should simply be encoded with its own code point. This happened
only with the revision of the Baudot-code (see below) by Donald Murray in 1901,
in which he introduced a specific white-space code. This principle has been
followed ever since.}

>>>>

Writing systems arise and develop in a complex mixture of cultural,
technological and practical pressures. They tend to be highly conservative, in
that people who have learned to read and write in a specific way -- however
impractical or tedious -- are mostly unwilling to change their habits. Writers
tend to resist spelling reforms. In all literate societies there exists a strong
socio-political mainstream that tries to force unification of writing (for
example by strongly enforcing ``right'' from ``wrong'' writing in schools).
However, there is also a large community of users who take as many liberties in
their writing as they can get away with.

@@

In the nineteenth century, electric telegraphy lead to a next innovation in
which written language characters were encoded by signals sent through a copper
wire. Originally, \textsc{bisignal codes} were used, consisting of two different
signals. For example, Carl Friedrich Gauss in 1833 used positive and negative
current \citep[282]{Mania2008}. More famous and influential, Samuel Morse in
1836 used long and short pulses. In those bisignal codes each character from the
written language was encoded with a different number of signals (between one and
five), so two different separators are needed: one between signals and one
between characters. For example, in Morse-code there is a short pause between
signals and a long pause between characters.\footnote{Actually, Morse-code also
includes an extra long pause between words. Interestingly, it took a long time
to consider the written word boundary -- using white-space -- as a bona-fide
character that should simply be encoded with its own code point. This happened
only with the revision of the Baudot-code (see below) by Donald Murray in 1901,
in which he introduced a specific white-space code. This principle has been
followed ever since.}

### 2018-06-04 book/chapters/pitfalls.tex
fixes en dashes throughout

In this chapter we describe some of the most common pitfalls that we have
encountered when using the Unicode Standard in our own work, or in discussion
with other linguists. This section is not meant as a criticism of the decisions
made by the Unicode Consortium; on the contrary, we aim to highlight where the
technological aspects of the Unicode Standard diverge from many users'
intuitions. What have sometimes been referred to as problems or inconsistencies
in the Unicode Standard are mostly due to legacy compatibility issues, which can
lead to unexpected behavior by linguists using the standard. However, there are
also some cases in which the Unicode Standard has made decisions that
theoretically could have been taken differently, but for some reason or another
(mostly very good reasons) were accepted as they are now. We call such behavior that
executes without error but does something different than the user
expected---often unknowingly---a \textsc{pitfall}.

@@

Homoglyphs are visually indistinguishable glyphs (or highly similar glyphs) that
have different code points in the Unicode Standard and thus different character
semantics. As a principle, the Unicode Standard does not specify how a character
appears visually on the page or the screen. So in most cases, a different
appearance is caused by the specific design of a font, or by user-settings like
size or boldface. Taking an example already discussed in
Section~\ref{pitfall-homoglyphs}, the following symbols <g {\large \textit{g}}
\textbf{g} {\fontspec{ArialMT} {\small g} \textit{g} \textbf{g}}> are different
glyphs of the same character, i.e.~they may be rendered differently depending on
the typography being used, but they all share the same code point (viz.
\textsc{latin small letter g} at \uni{0067}). In contrast, the symbols
<{\fontspec{EversonMono}AАΑᎪᗅᴀꓮ𐊠𝖠𝙰}> are all different code points,
although they look highly similar---in some cases even sharing exactly the same
glyph in some fonts. All these different A-like characters include the following
code points in the Unicode Standard:

>>>>

In this chapter we describe some of the most common pitfalls that we have
encountered when using the Unicode Standard in our own work, or in discussion
with other linguists. This section is not meant as a criticism of the decisions
made by the Unicode Consortium; on the contrary, we aim to highlight where the
technological aspects of the Unicode Standard diverge from many users'
intuitions. What have sometimes been referred to as problems or inconsistencies
in the Unicode Standard are mostly due to legacy compatibility issues, which can
lead to unexpected behavior by linguists using the standard. However, there are
also some cases in which the Unicode Standard has made decisions that
theoretically could have been taken differently, but for some reason or another
(mostly very good reasons) were accepted as they are now. We call such behavior that
executes without error but does something different than the user
expected -- often unknowingly -- a \textsc{pitfall}.

@@

Homoglyphs are visually indistinguishable glyphs (or highly similar glyphs) that
have different code points in the Unicode Standard and thus different character
semantics. As a principle, the Unicode Standard does not specify how a character
appears visually on the page or the screen. So in most cases, a different
appearance is caused by the specific design of a font, or by user-settings like
size or boldface. Taking an example already discussed in
Section~\ref{pitfall-homoglyphs}, the following symbols <g {\large \textit{g}}
\textbf{g} {\fontspec{ArialMT} {\small g} \textit{g} \textbf{g}}> are different
glyphs of the same character, i.e.~they may be rendered differently depending on
the typography being used, but they all share the same code point (viz.
\textsc{latin small letter g} at \uni{0067}). In contrast, the symbols
<{\fontspec{EversonMono}AАΑᎪᗅᴀꓮ𐊠𝖠𝙰}> are all different code points,
although they look highly similar -- in some cases even sharing exactly the same
glyph in some fonts. All these different A-like characters include the following
code points in the Unicode Standard:
